DRAFT: Direct-style process creation on Linux

### 2020-05-15 research/paper/paper.tex
paper stuff

\usepackage{hyperref}
\usepackage{float}
\usepackage{listings}
\usepackage{booktabs} % For formal tables

@@

% Title portion. Note the short title for running heads
\title[Direct-style process creation on Linux]{Direct-style process creation on Linux}

@@

In our implementation, a parent process calls the \texttt{clone} syscall
to create a child process which shares memory and other resources with the parent,
and which runs no user code.
The parent then uses \texttt{rsyscall}
to call arbitrary syscalls in the child process to initialize it;
for example, the parent might call \texttt{chdir} to change the child's current working directory.
After the process is fully initialized,
the parent can call \texttt{execve} in the child,
causing the process to start running an arbitrary executable
and detaching the child from the control of \texttt{rsyscall}.

A direct-style \texttt{clone} has all the advantages of direct-style process creation
over Linux's fork-style \texttt{fork} syscall and spawn-style \texttt{posix\_spawn} libc function.
Direct-style \texttt{clone} can be used correctly and performantly from any process.
Cross-process syscalls made with rsyscall individually report errors, as normal, and can be easily branched on.
Arbitrary processes involving complex namespaces and initialization can be created
using techniques that are impractical with \texttt{fork}.
The performance cost from high-level code varies, but the underlying primitives are substantially faster.

We have used direct-style process creation on Linux to do TODO and TODO,
and we've benchmarked our implementation as TODO.
Direct-style \texttt{clone} has minimal performance cost in normal situations,
and can outperform \texttt{fork} in some cases.
We have deployed direct-style process creation in production internally at Two Sigma.

@@

This paper is organized in several sections.
In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
and consider their advantages and disadvantages in more depth.
In section \ref{overview} we discuss the design of the API,
including our usage of \texttt{rsyscall}.
In section \ref{applications} we show examples of usage of direct-style process creation in Python on Linux,
including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
In section \ref{implementation}, we explore our implementation in depth,
and discuss several difficult process-creation related details of \texttt{rsyscall}'s implementation.
In section \ref{evaluation}, we evaluate, somehow (TODO).
In section \ref{future_work},
we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
In section \ref{conclusions}, we briefly and optimistically conclude.

@@

The most common complaint about fork is its poor performance \cite{forkroad}.
It copies many attributes about the parent process when creating the child process,
including setting up copy-on-write memory-mappings in the child process.
This becomes slower as the parent process has more memory-mappings,
eventually taking a significant amount of time.
This copying is required to robustly implement fork's model,
where the same program continues executing without changes in both the parent process and child process;
the copy of the program running in the child process might access any memory at any time.

Multi-threaded programs generally cannot use fork safely.
Typical Unix fork implementations duplicate only the thread calling fork from the parent process to the child process.
In a multi-threaded program, this can cause various issues;
for example, another thread might be holding a memory allocator lock at the time of the fork,
which in the child process will never be unlocked,
causing the child process to deadlock if it tries to allocate memory.
Some thread libraries provide partial mitigations for this issue,
but it's up to user code to make use of those mitigations.
In combination with fork's poor performance in large-memory programs,
a user of fork must think carefully
about the characteristics of the process from which fork is being called.
\subsection{Spawn-style}
In a spawn-style interface,
all the details about the new process are provided up-front as arguments to some function,
which creates the new process from a mostly-clean slate, initialized with the provided details.
The most well known spawn-style interfaces are Unix's \texttt{posix\_spawn} \cite{posix_spawn}
and Windows \texttt{CreateProcess} \cite{create_process};
the Unix shell also serves as a spawn-style process creation interface,
especially when used with chain-loading techniques.[fn:chainloading]

@@

Spawn-style interfaces lift the constraints on the calling process that fork-style interfaces impose.
Since spawn-style interfaces don't run user code in the new process during initialization,
there is no need to copy memory,
and there are no inherent issues when the interface is used from a multi-threaded program
(though operating systems which implement spawn-style interfaces on top of \texttt{fork} can still have bugs).

@@

Furthermore, spawn-style process creation
creates a new domain-specific language for specifying the attributes of the process.
Fork-style allows existing system calls,
with which programmers are already familiar,
to be used to set up the new process.
Spawn-style requires some form of declarative description of the new process,
which must be learned in additional to the normal system call interface.

Spawn-style process creation also does not allow for conditional logic during the setup;
if the setup of the new process encounters an error at some point,
the only option is to return from the entire spawn call with an error.
Such errors returned from spawn-style calls
are typically much less informative
than the errors returned by the syscalls which directly mutate the process attributes.
Other forms of conditional logic are also impossible in spawn-style;
one modification to the process cannot depend on the result of some other modification.
\subsection{Direct-style}
A few academic operating systems, such as KeyKOS \cite{keykos}, seL4 \cite{sel4}
and some others \cite{exokernel} \cite{fuschia} \cite{singularity},
use another style of process creation, which we refer to as "direct-style".

@@

Since everything happens directly from the parent process,
process initialization is compatible with a variety of techniques
which were otherwise incompatible with process creation.
For example, a program could provide a UI which directly customizes the new process,
or might initialize several processes at once and share data between them.

Most significantly,
since all operations are performed through syscalls called directly from the main program,
errors are indicated in the same way as any other syscall error.
This is unlike fork-style, which needs IPC to indicate errors,
and unlike spawn-style, which typically indicates errors at a very coarse-grained level.

@@

Like spawn-style interfaces,
direct-style interfaces have no constraints on the calling process.
No user code runs in the child process,
so, unlike fork-style interfaces,
direct-style interfaces can be used from multi-threaded or large-memory processes without issues.

@@

\texttt{rsyscall} provides this API for Linux.
The \texttt{rsyscall} project develops
language-specific, object-capability-model, low-abstraction libraries for the Linux syscall interface,
bypassing libc.
Unlike POSIX C libraries such as glibc or musl,
an rsyscall library is organized based on the object-capability model.
The capabilities to make syscalls in any given process are reified as language objects.
If the capability to make syscalls in a specific process is not passed (in some way) to a function,
then that function cannot make syscalls in that process.

@@

From the perspective of an program using rsyscall,
there are multiple processes which are under its control,
through multiple capabilities.
We use the term ``thread'' to refer to all such controlled processes,
including the main "thread" on which the program is running.
On Linux, the shared-memory "threads" provided by libraries such as \texttt{pthreads}
are implemented as processes,
and their lifetime and execution is completely controlled by a single program.
The same is mostly true of the controlled-process "threads" provided by rsyscall.
The most significant difference is that rsyscall "threads"
do not run their own code concurrently with the main program;
nevertheless rsyscall "threads" do provide the opportunity for parallel execution of system calls,
and so the terminology provides useful intuition.

@@

\begin{lstlisting}[float=*,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
child = local_thread.clone()
child.chdir("/dev")
child.execve("/bin/cat",
  ["cat", "./null"], local_thread.environ)
\end{lstlisting}
Listing \ref{basic} shows this in action in Python.
We create a new child thread using \texttt{clone} on \verb|local_thread|,
and receive back a capability to control it.
We pass no flags to use \texttt{clone}'s defaults
(which essentially match \texttt{fork}) to create the process.
We call \texttt{chdir} in the thread to change its working directory to \texttt{/dev}.
Then we call \texttt{execve},
passing the path of the \texttt{cat} executable,
the arguments to the new executable,
and the unchanged set of environment variables.
The name of the executable is passed as the first argument,
as is typical in Unix.
Calling \texttt{execve} consumes the capability,
and later use of this child thread capability will fail with an exception.

The API for direct-style process creation on Linux is not new;
it is the same Linux system call API that systems programmers are already familiar with.
The novelty is constrained to the newly explicit process capabilities;
once that is understood, the rest of the API is mundane,
and the representation of these programs is obvious in any language.

The familiar and mundane nature of the API is important not just for the learning curve,
but also for expressivity.
Since the API is simply normal Linux system calls,
anything that could be expressed with fork and syscalls
can also be expressed with direct-style clone.
\section{Applications}\label{applications}
In this section,
we'll discuss several more advanced uses of processes,
using more sophisticated Linux features,
and accompany them with Python examples using direct-style process creation.

% TODO this intro is all a bit muddled, and also not true.
These examples are all conventional applications of Linux features;
they are all possible using Linux's existing \texttt{fork} system call,
and in some cases, \verb|posix_spawn|,
with varying degrees of difficulty and workarounds.
But direct-style makes these easier to write;
we'll compare our implementations in this section to fork-style implementations in section \ref{evaluation}.
Furthermore, since the examples are using the Linux system call API directly,
they can be translated straightforwardly into any language with an \texttt{rsyscall} API;
Python is chosen simply for convenience.

Some of these examples showcase functionality that can be done today by specialized software and services.
Such software often allows use of one or two of these techniques, but prevents the use of others.

By improving the process creation interface,
all these techniques can be used together, flexibly,
to more closely meet the specific needs of individual real-world applications.
\subsection{Abstraction through file descriptor passing}\label{fd_abstraction}
In Linux (and Unix in general),
a program can use file descriptor inheritance
to provide file descriptors created by the parent process to child processes.
Put simply,
a new Linux child process receives copies of all file descriptors in the parent process.
When calling \texttt{execve}, these file descriptors are normally closed automatically,
but selected file descriptors can be preserved over \texttt{execve}
by disabling the \texttt{CLOEXEC} flag on individual file descriptors.
We will discuss file descriptor inheritance in more detail in section \label{fdtables},
as part of our discussion of resource management in the implementation of \texttt{rsyscall}.

Most resources in Linux are managed through file descriptors,
so this allows the parent process to pass a wide variety of resources to the child process.
An executable program can be provided a file descriptor,
which can correspond to any file in any filesystem,
or to a network connection\cite{ucspi},
or a pipe connected to another process,
without the process being aware of what specific resource it is accessing.
In this way, we immediately know "for free"\cite{theoremsforfree}
that the behavior of the program does not depend on the name of the file,
or the hostname to it which the socket is connected,
or the executable running in the process on the other end of the pipe.

\begin{lstlisting}[float=*,language=Python,label={fds},caption={Passing down FDs}]
db_fd = local_thread.open("/var/db/database.db", O.RDWR)
child = local_thread.clone()
child_fd = child.inherit_fd(db_fd)
child_fd.fcntl(F.SETFD, 0)
child.execv("/bin/fooserver",
  ["fooserver", "--database-fd", str(int(child_fd))])
\end{lstlisting}
In Listing \ref{fds},
we first open a file with read-write permission.
Then we create a child thread,
which inherits all file descriptors from its parent.
We indicate that we want to operate on the child's copy of the file descriptor with \verb|child.inherit_fd|.
\verb|inherit_fd| only updates resource-tracking bookkeeping in our program;
it makes no system calls.
Then we disable \texttt{CLOEXEC} on the child's file descriptor,
which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|.
Finally, we execute some executable,
passing the file descriptor number as an argument;
when the executable starts up, that file descriptor number will already be pointing to this file descriptor.
We call the \texttt{execv} wrapper rather than \texttt{execve} as in listing \ref{basic}
so that we don't need to pass the (unmodified) environment to every exec call in these examples.
\subsection{Non-shared-memory concurrency}
Processes run concurrently,
which allows exploiting the parallelism of the underlying hardware.
Since processes don't share memory,
they can provide a less complex parallel programming environment
than shared-memory thread-based approaches.

A popular parallel programming environment is the Unix shell,
which obtains its parallelism by running multiple processes connected via pipes.
The Unix shell has a relatively constrained form of parallel processing,
but it's also possible to create more complex webs of parallel processes,
where, for example, one process might take multiple inputs over multiple pipes,
or produce multiple outputs.

\begin{lstlisting}[float=*,language=Python,label={pipe},caption={Creating a parallel processing pipeline}]
def as_argument(child, fd):
  child_fd = child.inherit_fd(fd)
  child_fd.fcntl(F.SETFD, 0) # unset CLOEXEC
  return str(int(child_fd)) # turn fd number to string

audio_pipe = local_thread.pipe()
video_pipe = local_thread.pipe()
source = local.thread.clone()
source.execve('/bin/source', ['source',
  '--audio-output', as_argument(source, audio_pipe.write),
  '--video-output', as_argument(source, video_pipe.write)])
video_sink = local.thread.clone()
video_sink.execve('/bin/video_sink', ['video_sink',
  '--video-input', as_argument(video_sink, video_pipe.read)])
audio_sink = local.thread.clone()
audio_sink.execve('/bin/audio_sink', ['audio_sink',
  '--audio-input', as_argument(audio_sink, audio_pipe.read)])
\end{lstlisting}
In listing \ref{pipe},
we execute a few programs in parallel,
connected by pipes.
We first create two pipes,
then inherit them down to several processes using a helper function, \verb|as_argument|,
which uses \verb|inherit_fd| and \texttt{fcntl} as described in section \ref{fd_abstraction}.
We exec each program passing some file descriptors as arguments.
The whole system then runs in parallel,
as normal child processes.
\subsection{Customization without explicit support}
While, ideally, all resources would be passed by file descriptor,[fn:capsicum]
Unix has a number of resources that are global,
such as the pid space or the mount table,
and which cannot be changed on a per-process basis.

@@

Namespaces, like all process-local resources,
can be used to customize a program's behavior
in ways that were not anticipated by the author
by customizing the environment the process runs in.

\begin{lstlisting}[float=*,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
child = local.thread.clone(CLONE.NEWNS)
child.mount("/home/foo/custom_foo.conf", "/etc/foo.conf", "", MS.BIND, "")
child.execv('/bin/fooserver', ['fooserver'])
\end{lstlisting}
One classic form of customization enabled by namespaces
is overriding files at paths that were hardcoded into the program.
In listing \ref{mount},
we create a new child thread as normal,
but for the first time,
pass an argument to \texttt{clone}.
\verb|CLONE_NEWNS| causes clone to create the process in a new mount namespace,
which allows us to use the \texttt{mount} system call
to manipulate the filesystem tree in a way that only this process will see.
We call mount, passing \texttt{MS.BIND}, to make the file \verb|/home/foo/custom_foo.conf|
appear at the path \verb|/etc/foo.conf|;
this is known as a bind mount.
Then we execute some executable,
which will see the \verb|/etc/foo.conf| as we specified.
\subsection{Sandboxing}
At process creation time,
we can not only pass resources and customize the process's environment,
we can also deny the process access to resources that it otherwise gets by default.
This is a key part of creating a secure sandbox for potentially malicious code.

Historically, system calls such as \texttt{chroot} were the main mechanism for doing this;
today, Linux's namespaces system is the most important component.

\begin{lstlisting}[float=*,language=Python,label={unmount},caption={Unmount all and run executable via fexec}]
thread = local_thread.clone(CLONE.NEWNS)
executable_fd = child.open("/bin/static_executable", O.RDONLY)
db_fd = child.open("/var/db/database.db", O.RDWR)
child.umount("/", MNT.DETACH)
db_fd.fcntl(F.SETFD, 0)
child.fexec(executable_fd, ["static_executable", "--database-fd", str(int(db_fd))])
\end{lstlisting}
In listing \ref{unmount}
we create a new child thread,
again in a new mount namespace using \texttt{CLONE.NEWNS}.
We open the executable that we'll later exec in this child thread,
as well as another file as in listing \ref{fds},
since we won't be able to do exec or open files normally after the next step.
Then we use \texttt{umount},
passing \texttt{MNT.DETACH} to perform a recursive unmount of the entire filesystem tree,
removing it all from the view of this process.
As in listing \ref{fds}, we unset the \texttt{CLOEXEC} flag from the database file descriptor,
so that it can be inherited across \texttt{exec};
we don't need to call \verb|inherit_fd| since, in this case, the database fd was opened directly from the child.

@@

Even if the process needs additional resources,
those can be explicitly passed down through file descriptor passing,
as we do here with the database file descriptor.
This is the essence of capability-based security,
which Linux can increasingly closely approximate.
\subsection{Nested clone and pid namespaces}
Process-local resource can also be used to control the lifetime of resources used by that process.
Many Linux resources are not automatically cleaned up on process exit;
a poorly coded program may allocate global resources
and not ensure that they are cleaned up,
leaving behind unuseds litter on the system.

@@

We can use a pid namespace, in conjunction with nested cloning, to solve this issue.
The lifetime of all processes in a pid namespace is tied to the first process created in it,
the init process.
When the init process dies,
all other processes in the pid namespace are destroyed.

\begin{lstlisting}[float=*,language=Python,label={pidns},caption={Nested clone and pid namespace}]
init = local_thread.clone(CLONE.NEWPID)
grandchild = init.clone()
grandchild.execv('/bin/fooserver', ['fooserver'])
\end{lstlisting}
In listing \ref{pidns},
we create a child process which we call \texttt{init},
passing \verb|CLONE_NEWPID| to create it in a new pid namespace.
To create another process in the pid namespace,
we clone again, this time from \texttt{init};
note that this is the first time we've cloned from one of our child processes.
We exec in the grandchild,
and we can monitor the grandchild process from \texttt{init},
just as we would monitor \texttt{init} from its parent process.

@@

In either case, when \texttt{init} dies,
the pid namespace will be destroyed,
and \texttt{grandchild} will be cleaned up.
\subsection{Shared file descriptor tables}
A library can create a child process to perform a useful service.
This can sometimes be done more easily when the child process shares file descriptors with the parent.

>>>>

\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{float}
\usepackage{listings}
\usepackage{booktabs} % For formal tables

@@

% Title portion. Note the short title for running heads
\title[DRAFT: Direct-style process creation on Linux]{DRAFT: Direct-style process creation on Linux}

@@

% TODO this sounds low-content and boring.
% we should probably remove it?
% In our implementation, a parent process calls the \texttt{clone} syscall
% to create a child process which shares memory and other resources with the parent,
% and which runs no user code.
% The parent then uses \texttt{rsyscall}
% to call arbitrary syscalls in the child process to initialize it;
% for example, the parent might call \texttt{chdir} to change the child's current working directory.
% After the process is fully initialized,
% the parent can call \texttt{execve} in the child,
% causing the process to start running an arbitrary executable
% and detaching the child from the control of \texttt{rsyscall}.

Our direct-style process creation implementation uses \texttt{clone} to create a child process in an embryonic state
which can then be manipulated from user code using syscalls with \texttt{rsyscall}.
A direct-style \texttt{clone} has all the advantages of direct-style process creation
over Linux's fork-style \texttt{fork} syscall and spawn-style \texttt{posix\_spawn} libc function.
Direct-style \texttt{clone} can be used correctly and performantly from any process.
Cross-process syscalls made with rsyscall individually report errors, as normal, and can be easily branched on.
Arbitrary processes involving complex namespaces and initialization can be created
using techniques that are impractical with \texttt{fork}.
The performance cost from high-level code varies, but the underlying primitives are substantially faster.

We've implemented a variety of useful applications using direct-style process creation on Linux,
which can be expressed more concisely and more clearly than with fork-style,
which we'll showcase in brief in this paper.
Direct-style \texttt{clone} has minimal performance cost in normal situations,
and can outperform \texttt{fork} in some cases.
We've benchmarked our Python implementation of direct-style \texttt{clone}
as comparable to the Python standard library's process creation mechanism.
Finally, we have deployed direct-style process creation in production internally at Two Sigma.

@@

This paper is organized in several sections.
In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
and consider their advantages and disadvantages in more depth.
In section \ref{overview} we discuss the design of the API,
including our usage of \texttt{rsyscall}.
In section \ref{applications} we show examples of usage of direct-style process creation in Python on Linux,
including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
In section \ref{implementation}, we explore our implementation in depth,
and discuss several difficult process-creation related details of \texttt{rsyscall}'s implementation.
In section \ref{evaluation}, we evaluate, somehow (TODO).
In section \ref{related_work},
we examine other work which also touches on the topics covered by this paper.
In section \ref{future_work},
we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
In section \ref{conclusions}, we briefly and optimistically conclude.

@@

Fork can have poor performance when called from processes with many memory mappings.\cite{forkroad}
Fork copies many attributes about the parent process when creating the child process,
including setting up copy-on-write memory mappings in the child process.
This becomes slower as the parent process has more memory mappings,
eventually taking a significant amount of time.
This copying is required to robustly implement fork's model,
where the same program continues executing without changes in both the parent process and child process;
the copy of the program running in the child process might access any memory at any time.

Multi-threaded programs generally cannot use fork safely.
Typical Unix fork implementations duplicate only the thread calling fork from the parent process to the child process.
In a multi-threaded program, this can cause various issues;
for example, another thread might be holding a memory allocator lock at the time of the fork,
which in the child process will never be unlocked,
causing the child process to deadlock if it tries to allocate memory.
Some thread libraries provide partial mitigations for this issue,
but it's up to user code to make use of those mitigations.\cite{pthread_atfork}
In combination with fork's poor performance in large-memory programs,
a user of fork must think carefully
about the characteristics of the process from which fork is being called.
\subsection{Spawn-style}
In a spawn-style interface,
details about the new process are provided up-front as arguments to some function
which creates the new process all at once.
The most well known spawn-style interfaces are Unix's \texttt{posix\_spawn} \cite{posix_spawn}
and Windows \texttt{CreateProcess} \cite{create_process};
the Unix shell also serves as a spawn-style process creation interface,
especially when used with Bernstein chaining.\cite{chainloading}

@@

Spawn-style interfaces lift the constraints on the calling process that fork-style interfaces impose.
Spawn-style interfaces don't run user code in the new process during initialization.
This means there is no need to copy memory,
so performance is good when a spawn-style interface is used from a process with many memory mappings.
Likewise, there are no inherent issues when a spawn-style interface is used from a multi-threaded program,
though operating systems which implement spawn-style interfaces on top of \texttt{fork} can still have threading bugs.

@@

Spawn-style process creation
essentially creates a new domain-specific language for specifying the attributes of the process.
Fork-style allows existing system calls,
with which programmers are already familiar,
to be used to set up the new process.
Spawn-style requires some form of declarative description of the new process,
which must be learned in additional to the normal system call interface.

Spawn-style process creation also does not allow for conditional logic during the setup.
If the setup of the new process encounters an error at some point,
the only option is to return from the entire spawn call with an error.
Such errors returned from spawn-style calls
are typically much less informative
than the errors returned by the syscalls which directly mutate the process attributes.
Other forms of conditional logic are also impossible in spawn-style;
one modification to the process cannot depend on the result of some other modification.
\subsection{Direct-style}
A few academic operating systems, such as KeyKOS \cite{keykos}, seL4 \cite{sel4}
and some others \cite{exokernel} \cite{fuschia} \cite{singularity},
use another style of process creation, which we refer to as "direct-style".

@@

Since everything happens directly from the parent process,
direct-style process creation is compatible with a variety of techniques
which are difficult with fork-style or spawn-style.
For example, a REPL could be used to interactively create the new process,
or a program might initialize several processes at once and share data between them.

Significantly,
since all operations are performed through syscalls called directly from the main program,
errors are indicated in the same way as any other syscall error.
This is unlike fork-style, which needs IPC to indicate errors,
and unlike spawn-style, which typically indicates errors at a very coarse-grained level.

@@

Like spawn-style interfaces,
direct-style interfaces have no constraints on the calling process.
Also like spawn-style interface,
no user code runs in the child process, so
direct-style interfaces can be used from multi-threaded or large-memory processes without issues.

@@

\texttt{rsyscall} provides this API for Linux.
The \texttt{rsyscall} project develops
language-specific, object-capability-model, low-abstraction libraries for the Linux syscall interface,
bypassing libc.

Unlike POSIX C libraries such as glibc or musl,
an rsyscall library is organized based on the object-capability model.
The capabilities to make syscalls in any given process are reified as language objects.
If the capability to make syscalls in a specific process is not passed (in some way) to a function,
then that function cannot make syscalls in that process.

@@

We use the term ``thread'' to refer to the processes under the control of rsyscall,
including the ``local'' process.
On Linux, the shared-memory ``threads'' provided by libraries such as \texttt{pthreads}
are implemented as processes,
and their lifetime and execution is completely controlled by a single program.
The same is mostly true of the controlled-process ``threads'' provided by rsyscall.
The most significant difference is that rsyscall ``threads''
do not run user code concurrently with the main program;
nevertheless rsyscall ``threads'' do provide the opportunity for concurrent execution of system calls,
and so the terminology provides useful intuition.

@@

\lstinputlisting[float=*,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
{examples/basic/direct.py}
Listing \ref{basic} shows this in action in Python.
We create a new child thread using \texttt{clone} on \verb|local_thread|,
and receive back a capability to control it.
We pass no flags to use \texttt{clone}'s defaults
(which essentially match \texttt{fork}) to create the process.
We call \texttt{chdir} in the thread to change its working directory to \texttt{/dev}.
Then we call \texttt{execve},
passing the path of the \texttt{cat} executable,
the arguments to the new executable,
and the unchanged set of environment variables.
The name of the executable is passed as the first argument,
as is typical in Unix.
Calling \texttt{execve} consumes the capability,
and later use of this child thread capability will fail with an exception.

The behavior of system calls made on a remote process through \texttt{rsyscall}
is identical to the behavior of system calls made implicitly on the current process through libc.
This is an API that systems programmers are already familiar with,
and which is already thoroughly documented through Linux system call manpages.
The novelty is constrained to the newly explicit process capabilities;
once that is understood, the rest of the API is mundane,
and the representation of these programs is obvious in any language.

The familiar and mundane nature of the API is important not just for the learning curve,
but also for expressivity.
Since the API is simply normal Linux system calls,
direct-style clone is at least as expressive as fork and syscalls;
and, in fact, as we'll see in section \ref{applications},
is significantly more expressive.
\section{Applications}\label{applications}
In this section,
we'll discuss several more advanced uses of processes,
using more sophisticated Linux features,
and accompany them with Python examples using direct-style process creation.

All of these examples use only existing Linux system calls and features.
Most are possible, though difficult to implement,
with Linux's existing \texttt{fork} system call,
and in some cases, \verb|posix_spawn|.

Direct-style makes these examples substantially easier to write;
we'll compare our implementations in this section to fork-style implementations in section \ref{evaluation}.
Furthermore, since the examples are using the Linux system call API directly,
they can be translated straightforwardly into any language with an \texttt{rsyscall} API;
Python is chosen simply for convenience.

Some of these examples showcase functionality that can be done today by specialized software and services.
Such software often allows use of one or two of these techniques, but prevents the use of others.
For example, a shell allows the creation of pipelines and container systems like Docker allow sandboxing,
but the two are difficult to combine.

By improving the process creation interface,
all these techniques can be used together, flexibly,
to more closely meet the specific needs of individual real-world applications.

For concision, we assume that we are running with sufficient permissions in these examples,
but these examples can all be run without privileges with appropriate use of \texttt{CLONE.NEWUSER}.
\subsection{Abstraction through file descriptor passing}\label{fd_abstraction}
In Linux (and Unix in general),
a program can use file descriptor inheritance
to provide file descriptors created by the parent process to child processes.
Put simply,
a new Linux child process receives copies of all file descriptors in the parent process.
When calling \texttt{execve}, these file descriptors are normally closed automatically,
but selected file descriptors can be preserved over \texttt{execve}
by disabling the \texttt{CLOEXEC} flag on individual file descriptors.
We will discuss file descriptor inheritance in more detail in section \label{fdtables},
as part of our discussion of resource management in the implementation of \texttt{rsyscall}.

Most resources in Linux are managed through file descriptors,
so this allows the parent process to pass a wide variety of resources to the child process.
An executable program can be provided a file descriptor,
which can correspond to any file in any filesystem,
or to a network connection\cite{ucspi},
or a pipe connected to another process,
or a variety of other resources,
without the process being aware of what specific resource it is accessing.
In this way, we immediately know "for free"\cite{theoremsforfree}
that the behavior of the program does not depend on the name of the file,
or the hostname to it which the socket is connected,
or the executable running in the process on the other end of the pipe,
or other abstracted-away details.

\lstinputlisting[float=*,language=Python,label={fds},caption={Passing down FDs}]
{examples/fds/direct.py}
In Listing \ref{fds},
we first open a file with read-write permission.
Then we create a child thread,
which inherits all file descriptors from its parent.
We indicate that we want to operate on the child's copy of the file descriptor with \verb|child.inherit_fd|.
\verb|inherit_fd| only updates resource-tracking bookkeeping in our program,
it makes no system calls;
the file descriptor was already copied at process creation time.
Then we disable \texttt{CLOEXEC} on the child's file descriptor,
which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|.
Finally, we execute some executable,
passing the file descriptor number as an argument;
when the executable starts up, that file descriptor number will already be pointing to this file descriptor.
We call the \texttt{execv} wrapper rather than \texttt{execve} as in listing \ref{basic}
so that we don't need to pass the (unmodified) environment to every exec call in these examples.
\subsection{Non-shared-memory concurrency}
Processes run concurrently,
which allows exploiting the parallelism of the underlying hardware.
Since processes don't share memory,
they can provide a less complex parallel programming environment
than shared-memory thread-based approaches.

One of the most popular parallel programming environments is the Unix shell,
which obtains its parallelism by running multiple processes connected via pipes.
The Unix shell has a relatively constrained form of parallel processing,
but it's also possible to create more complex webs of parallel processes,
where, for example, one process might take multiple inputs over multiple pipes,
or produce multiple outputs.

\lstinputlisting[float=*,language=Python,label={pipe},caption={Creating a parallel processing pipeline}]
{examples/pipe/direct.py}
In listing \ref{pipe},
we execute a few programs in parallel,
connected by pipes.
We first create two pipes,
then inherit them down to several processes using a helper function, \verb|as_argument|,
which uses \verb|inherit_fd| and \texttt{fcntl} as described in section \ref{fd_abstraction}.
We exec each program passing some file descriptors as arguments.
The whole system then runs in parallel,
as normal child processes.
\subsection{Customization without explicit support}
While, ideally, all resources would be passed by file descriptor,\cite{capsicum}
Unix has a number of resources that are global,
such as the pid space or the mount table,
and which cannot be changed on a per-process basis.

@@

Namespaces, like all process-local resources,
can be used to customize a program's behavior
in ways that were not anticipated by the author
by customizing the environment the process runs in.
Other system calls such as \texttt{chroot} can also be used for this.

\lstinputlisting[float=*,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
{examples/mount/direct.py}
One classic form of customization enabled by namespaces
is overriding files at paths that were hardcoded into a program we can't change,
without changing those files for the rest of the system.
In listing \ref{mount},
we create a new child thread as normal,
but for the first time,
pass an argument to \texttt{clone}.

\verb|CLONE_NEWNS| causes clone to create the process in a new mount namespace,
which allows us to manipulate the filesystem tree in a way that only this process will see
using the \texttt{mount} system call.
We call \texttt{mount}, passing \texttt{MS.BIND}, to make the file \verb|/home/foo/custom_foo.conf|
appear at the path \verb|/etc/foo.conf|.
This is known as a bind mount.
Then we execute some executable,
which, when it opens \verb|/etc/foo.conf|, will see the contents of our \verb|custom_foo.conf|.
\subsection{Sandboxing}
At process creation time,
we can not only pass resources and customize the process's environment,
we can also deny the process access to resources that it otherwise gets by default.
This is a key part of creating a secure sandbox for potentially malicious code.

\lstinputlisting[float=*,language=Python,label={unmount},caption={Unmount all and run executable via fexec}]
{examples/unmount/direct.py}
In listing \ref{unmount}
we create a new child thread,
again in a new mount namespace using \texttt{CLONE.NEWNS}.
We open the executable that we'll later exec in this child thread,
as well as another file as in listing \ref{fds},
since we won't be able to do exec or open files normally after the next step.
Then we use \texttt{umount},
passing \texttt{MNT.DETACH} to perform a recursive unmount of the entire filesystem tree,
removing it all from the view of this process.
As in listing \ref{fds}, we unset the \texttt{CLOEXEC} flag from the database file descriptor,
so that it can be inherited across \texttt{exec};
we don't need to call \verb|inherit_fd| since, in this case, the database fd was opened directly from the child.

@@

Even if the process needs additional resources,
those can be explicitly passed down through file descriptor passing,
as we do here with the database file descriptor.
This is the essence of capability-based security,
which Linux can increasingly closely approximate.
\subsection{Nested clone and pid namespaces}
Process-local resources can also be used to control the lifetime of resources used by that process.
Many Linux resources are not automatically cleaned up on process exit;
a poorly coded program may allocate global resources
and not ensure that they are cleaned up,
leaving behind unused garbage on the system.

@@

We can use a pid namespace to solve this issue.
The lifetime of all processes in a pid namespace is tied to the first process created in it,
the init process.
When the init process dies,
all other processes in the pid namespace are destroyed.

\lstinputlisting[float=*,language=Python,label={pidns},caption={Nested clone and pid namespace}]
{examples/pidns/direct.py}
In listing \ref{pidns},
we create a child process which we call \texttt{init},
passing \verb|CLONE_NEWPID| to create it in a new pid namespace.

To create another process in the pid namespace,
we clone again, this time from \texttt{init}.
This is the first example in which we've cloned from one of our child threads;
as normal, this produces a new process, a grandchild,
which can be used exactly like a direct child thread.
We exec in the grandchild,
and we can monitor the grandchild process from \texttt{init},
just as we would monitor \texttt{init} from its parent process.

@@

In either case, when \texttt{init} dies,
the pid namespace will be destroyed,
and \texttt{grandchild} will be cleaned up.
\subsection{Shared file descriptor tables}\label{shared_fd_table}
A library can create a child process to perform a useful service.
This can sometimes be done more easily when the child process shares file descriptors with the parent.

\lstinputlisting[float=*,language=Python,label={fuse},caption={Shared file descriptor tables}]
{examples/fuse/direct.py}
In listing \ref{fuse},
we first create a child process, \verb|ns_child|, in a new mount namespace with \verb|CLONE_NEWNS|,
and for the first time, also pass \verb|CLONE_FILES|,
which causes the file descriptor table to be shared between the parent process and the child process.
We create a grandchild off of \verb|ns_child|,
and use it to exec a FUSE filesystem,
which will appear only in the mount namespace of \verb|ns_child| and its descendents.
Then we can open FUSE files in \verb|ns_child|
and use those files in the parent process,
through the shared file descriptor table.

File descriptor passing over a Unix domain socket
can allow similar behavior without sharing the file descriptor table,
but is substantially more complex for the user.
Nevertheless, for fork-style and spawn-style process creation,
file descriptor passing over a Unix domain socket
is the only option to do something like this.
\section{Implementation}\label{implementation}
\subsection{Basics about rsyscall}
In this section, we'll give a brief overview of the implementation of rsyscall,
and then focus on implementation issues specific to process creation.

rsyscall can be conceptually divided in two parts:
the basic cross-process syscall primitive,
and a language-specific library built on top.
The Python language-specific library has already been demonstrated above.
Such libraries only need to be able to call syscalls and explicitly specify a process in some way;
they are, for the most part, agnostic to how the cross-process syscall is implemented.
